---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# morrowplots

<!-- badges: start -->
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/TBD)](https://cran.r-project.org/package=morrowplots)
[![R-CMD-check](https://github.com/SandiCal/morrowplots/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/SandiCal/morrowplots/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Use morrowplots to load, explore, and analyze nearly 150 years of ag data from the historic [Morrow Plots](https://aces.illinois.edu/research/history/morrow-plots) at the [University of Illinois Urbana-Champaign](https://illinois.edu/). Data collection and formats have evolved over time, and this dataset was created to aggregate decades of tabular data (previously in various physical and digital forms) into a format compatible with the R programming language. Vignettes provide mini lessons on the history of the Morrow Plots and example code for exploration of the data.

## Installation

(Update on CRAN release!) You can install the **stable** version from
[CRAN](https://cran.r-project.org/package=morrowplots).

```r
install.packages("morrowplots")
```

You can install the **development** version [Github](https://github.com/SandiCal/morrowplots):

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("SandiCal/morrowplots")
```

## Structure

Data is structured as a dataframe in the [Tidy Data](https://vita.had.co.nz/papers/tidy-data.pdf) format so that each
variable has its own column, each observation has its own row, and each
value has its own cell.
```{r, layout="l-body-outset"}
library(morrowplots)
library(knitr)
kable(head(morrowplots))
```

## Variables

The codebook that describes all variables is available in the help documentation.

```{r}
help(morrowplots)
```

```{r, echo = F}
helpfile <- utils:::.getHelpFile(help(morrowplots))
outfile <- tempfile(fileext = ".html")
tools:::Rd2HTML(helpfile, out =outfile)
rawHTML <- paste(readLines(outfile), collapse="\n")
knitr::asis_output(htmltools::htmlPreserve(rawHTML))
```
