---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# morrowplots

<!-- badges: start -->
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/TBD)](https://cran.r-project.org/package=morrowplots)
<!-- badges: end -->

Use morrowplots to load, explore, and analyze nearly 150 years of ag data from the historic [Morrow Plots](https://aces.illinois.edu/research/history/morrow-plots) at the [University of Illinois Urbana-Champaign](https://illinois.edu/). Data collection and formats have evolved over time, and this dataset was created to aggregate decades of tabular data (previously in various physical and digital forms) into a format compatible with the R programming language. Vignettes provide mini lessons on the history of the Morrow Plots and example code for exploration of the data.

## Installation

(Update on CRAN release!) You can install the **stable** version from
[CRAN](https://cran.r-project.org/package=morrowplots).

```r
install.packages("morrowplots")
```

You can install the **development** version [Github](https://github.com/SandiCal/morrowplots):

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("SandiCal/morrowplots")
```

## Structure

Data is structured as a dataframe containing 26 variables and 3216 rows in the [Tidy Data](https://vita.had.co.nz/papers/tidy-data.pdf) format. 
```{r, layout="l-body-outset"}
library(morrowplots)
library(knitr)
kable(head(morrowplots))
```

## Variables

You can view the codebook that describes all variables in the help.

```{r}
help(morrowplots)
```

## - 
## - 
## - initial helper text and other notes
## - *** to delete later ***

Data Package README examples:

BEST - 
* https://github.com/forestgeo/fgeo.x - repo  
* https://forestgeo.github.io/fgeo.x/ - page

Others - 
* https://github.com/hadley/babynames
* https://github.com/robjhyndman/cricketdata

## - 
## - 

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.

<iframe width="560" height="315" src="https://www.youtube.com/embed/hsipxaNGnbc?si=2xIMfY8ndPYl_zSZ" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen> </iframe>
